---
title: "Comprehensive Health and Well-Being Study Among Singaporean Seniors"
author: "Krystle Teo"
date: "2024-08-08"
output: html_document
---

## Introduction:
This script is designed to analyze the impact of a 10-week holistic health program on various health constructs among seniors in Singapore. The script performs data pre-processing, creates an SQLite database, executes SQL queries,and conducts statistical analyses to explore the relationships between different health constructs/ outcome measures. The results are visualized and documented for further interpretation.

## Instructions:
- Ensure that the necessary R packages listed below are installed.
- Set the working directory to the folder containing your data files and SQLite database using setwd().
- Run the entire script to preprocess data, create the SQLite database, and perform analyses.

## Set-up environment and load packages needed
First load the necessary libraries and scripts.

### Set the working directory to the location of your data files
```{r}
setwd("/Users/krystle/Documents/PSYCH755/final_project")  # <<<< CHANGE THIS TO YOUR OWN DIRECTORY PATH
```

### Set up conflict policies 
```{r}
options(conflicts.policy = "depends.ok")
devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/fun_ml.R?raw=true")
tidymodels_conflictRules()

```

### Load packages needed
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
library(RSQLite)
library(dplyr)
library(DBI)
library(tidyverse)
```

### Source scripts 
```{r}
devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/fun_eda.R?raw=true")
devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/fun_plots.R?raw=true")
```

## Data pre-processing
In this section, load the CSV files containing the raw data and perform necessary data cleaning operations.

### Load csv files
```{r}
d1 <- read_csv("2023_SHAPESG_Data_TP1.csv")
d2 <- read_csv("2023_SHAPESG_Data_TP2.csv")
```

### Pre-processing and cleaning of datasets
```{r}
d1_clean <- d1 |> 
  mutate(across(where(is.character), ~ na_if(., "#DIV/0!"))) # Replace "#DIV/0!" with NA in all character columns

# Check for list of shapesg_id with all values as NA
na_rows_d1 <- d1_clean |> 
  filter(if_all(-SHAPESG_ID, is.na)) |> 
  pull(SHAPESG_ID)

# Remove NA rows and clean column names
d1_clean <- d1_clean |> 
  filter(!SHAPESG_ID %in% na_rows_d1) |> 
  janitor::clean_names() 

# Repeat same cleaning process for d2 df
d2_clean <- d2 |> 
  mutate(across(where(is.character), ~ na_if(., "#DIV/0!"))) # replace "#DIV/0!" with NA in all character columns

na_rows_d2 <- d2_clean |> 
  filter(if_all(-SHAPESG_ID, is.na)) |> 
  pull(SHAPESG_ID)

d2_clean <- d2_clean |> 
  filter(!SHAPESG_ID %in% na_rows_d2) |> 
  janitor::clean_names()
```
Both datasets were first modified to replace instances of "#DIV/0!" with NA in all character columns, effectively handling invalid entries. Following this, rows where all columns (except SHAPESG_ID) contained NA values were identified and removed, ensuring that only rows with meaningful data were retained. The column names were then standardized using janitor::clean_names(), making them more consistent and easier to work with. 

## Create new SQLite database using R
The database will be created in the current working directory.

### Connect to SQLite database
```{r}
db_file <- "wellbeing_database.sqlite"  
conn <- dbConnect(RSQLite::SQLite(), dbname = db_file)
```

### Write the cleaned data files to the SQLite database
```{r}
dbWriteTable(conn, "tp1", d1_clean, overwrite = TRUE)
dbWriteTable(conn, "tp2", d2_clean, overwrite = TRUE)
```

### List tables to confirm
```{r}
dbListTables(conn)

```

## Querying data
Now we can start querying some data from the tables.

### Querying 'tp1' table
```{r}
tp1 <- tbl(conn, "tp1")

count_tp1 <- tp1 |> 
  summarise(count_tp1 = n_distinct(shapesg_id)) |> 
  print()

show_query(count_tp1)
```

### Querying 'tp2' table
```{r}
tp2 <- tbl(conn, "tp2")

count_tp2 <- tp2 |> 
  summarise(count_tp2 = n_distinct(shapesg_id)) |> 
  print()

show_query(count_tp2)

```

### Querying sleep
Let us analyze sleep quality before and after the program.
```{r}
# Queries to sum sleep scores
query_sleep_sum_tp1 <- "
  SELECT shapesg_id, 
         (gsq1_01 + gsq1_02 + gsq1_03 + gsq1_04 + gsq1_05 + gsq1_06) AS total_sleep_score_tp1
  FROM tp1
"
  
sleep_scores_tp1 <- dbGetQuery(conn, query_sleep_sum_tp1)
sleep_scores_tp1 

query_sleep_sum_tp2 <- "
  SELECT shapesg_id, 
         (gsq2_01 + gsq2_02 + gsq2_03 + gsq2_04 + gsq2_05 + gsq2_06) AS total_sleep_score_tp2
  FROM tp2
"
  
sleep_scores_tp2 <- dbGetQuery(conn, query_sleep_sum_tp2)
sleep_scores_tp2

# Merge results
sleep_scores <- merge(sleep_scores_tp1, sleep_scores_tp2, by = "shapesg_id", all = TRUE)

# Calculate summary statistics for sleep scores in tp1 and tp2 
sleep_summary_tp1 <- summary(sleep_scores$total_sleep_score_tp1)
sleep_summary_tp2 <- summary(sleep_scores$total_sleep_score_tp2)

# Create a union of all possible statistic names
sleep_all_stats <- union(names(sleep_summary_tp1), names(sleep_summary_tp2))

# Align summaries by filling missing statistics with NA
sleep_summary_tp1 <- as.vector(sleep_summary_tp1[sleep_all_stats])
sleep_summary_tp2 <- as.vector(sleep_summary_tp2[sleep_all_stats])

# Create a data frame with the aligned summaries
sleep_summary_stats <- data.frame(
  Statistic = sleep_all_stats,
  TP1 = sleep_summary_tp1,
  TP2 = sleep_summary_tp2
) |> 
  print()

```
*Note that lower score indicates better sleep quality.*

```{r}
# Reshape data for ggplot
sleep_scores_long <- sleep_scores %>%
  pivot_longer(cols = starts_with("total_sleep_score"), 
               names_to = "time_point", 
               values_to = "sleep_score")

# Create boxplot
sleep_change <- 
  ggplot(sleep_scores_long, aes(x = time_point, y = sleep_score, fill = time_point)) +
  geom_boxplot() +
  labs(title = "Comparison of Sleep Scores Between TP1 and TP2",
       x = "Time Point",
       y = "Total Sleep Score") +
  theme_minimal()

sleep_change
```

### Querying BMI
```{r}
# Query BMI from tp1 and tp2, joining on shapesg_id
query_BMI <- "
  SELECT tp1.shapesg_id, 
         tp1.bmi1 AS bmi_tp1,
         tp2.bmi2 AS bmi_tp2
  FROM tp1
  JOIN tp2
  ON tp1.shapesg_id = tp2.shapesg_id
"
bmi_data <- dbGetQuery(conn, query_BMI)
bmi_data
```

```{r}
# Ensure that bmi_tp1 and bmi_tp2 are numeric
bmi_data$bmi_tp1 <- as.numeric(bmi_data$bmi_tp1)
bmi_data$bmi_tp2 <- as.numeric(bmi_data$bmi_tp2)

# Calculate mean BMI for each time point
mean_bmi_tp1 <- mean(bmi_data$bmi_tp1, na.rm = TRUE) |> print()
mean_bmi_tp2 <- mean(bmi_data$bmi_tp2, na.rm = TRUE) |> print()

```

```{r}
# Reshape data for proper plotting
bmi_data_long <- bmi_data %>%
  pivot_longer(cols = c(bmi_tp1, bmi_tp2), names_to = "time_point", values_to = "BMI")

# Plot side-by-side boxplots for TP1 and TP2
bmi_change <- 
  ggplot(bmi_data_long, aes(x = time_point, y = BMI, fill = time_point)) +
  geom_boxplot() +
  labs(title = "Comparison of BMI at TP1 and TP2",
       x = "Time Points",
       y = "BMI") +
  theme_minimal()
bmi_change
```

### Querying waist-hip ratio (WHR)
```{r}
# Query WHR from tp1 and tp2, joining on shapesg_id
query_WHR <- "
  SELECT tp1.shapesg_id, 
         tp1.whr1 AS whr_tp1,
         tp2.whr2 AS whr_tp2
  FROM tp1
  JOIN tp2
  ON tp1.shapesg_id = tp2.shapesg_id
"
whr_data <- dbGetQuery(conn, query_WHR)

# Print the queried data
print(whr_data)
```


```{r}
# Convert WHR columns to numeric 
whr_data$whr_tp1 <- as.numeric(whr_data$whr_tp1)
whr_data$whr_tp2 <- as.numeric(whr_data$whr_tp2)


# Calculate mean WHR for each time point
mean_whr_tp1 <- mean(whr_data$whr_tp1, na.rm = TRUE) |> print()
mean_whr_tp2 <- mean(whr_data$whr_tp2, na.rm = TRUE) |> print()

# Perform a paired t-test to compare WHR between TP1 and TP2
t_test_whr <- t.test(whr_data$whr_tp1, whr_data$whr_tp2, paired = TRUE)
print(t_test_whr)

# Visualize the comparison using a boxplot
whr_data_long <- whr_data %>%
  pivot_longer(cols = c(whr_tp1, whr_tp2), names_to = "time_point", values_to = "WHR")

whr_change <- 
  ggplot(whr_data_long, aes(x = time_point, y = WHR, fill = time_point)) +
  geom_boxplot() +
  labs(title = "Comparison of WHR at TP1 and TP2",
       x = "Time Points",
       y = "Waist-Hip Ratio (WHR)") +
  theme_minimal()

whr_change
```
### Querying waist-hip ratio (WHR) by gender
```{r}
query_gender <- "
  SELECT shapesg_id, gender
  FROM tp1
"

# Retrieve gender information
gender_data <- dbGetQuery(conn, query_gender)

# Join gender data with existing whr_data
whr_data <- merge(whr_data, gender_data, by = "shapesg_id", all.x = TRUE)

# View the updated data with gender included
print(whr_data)

```

#### Investigate gender differences 
```{r}
whr_data$gender <- as.factor(whr_data$gender)  # Convert gender to a factor for easier analysis

# Split the data by gender
whr_data_male <- whr_data |>  filter(gender == "1")  # Male
whr_data_female <- whr_data |>  filter(gender == "2")  # Female

# Calculate mean WHR for each gender and time point
mean_whr_tp1_male <- mean(whr_data_male$whr_tp1, na.rm = TRUE) |> print()
mean_whr_tp2_male <- mean(whr_data_male$whr_tp2, na.rm = TRUE) |> print()
mean_whr_tp1_female <- mean(whr_data_female$whr_tp1, na.rm = TRUE) |> print()
mean_whr_tp2_female <- mean(whr_data_female$whr_tp2, na.rm = TRUE) |> print()

# Perform paired t-tests by gender
t_test_whr_male <- t.test(whr_data_male$whr_tp1, whr_data_male$whr_tp2, paired = TRUE)
t_test_whr_female <- t.test(whr_data_female$whr_tp1, whr_data_female$whr_tp2, paired = TRUE)

t_test_whr_male
t_test_whr_female

# Visualize the comparison using a boxplot, grouped by gender
# Reshape data for proper plotting
whr_data_long <- whr_data %>%
  pivot_longer(cols = c(whr_tp1, whr_tp2), names_to = "time_point", values_to = "WHR")

whr_change_gender <- 
  ggplot(whr_data_long, aes(x = time_point, y = WHR, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~ gender, scales = "free", labeller = as_labeller(c(`1` = "Male", `2` = "Female"))) +
  labs(title = "Comparison of WHR at TP1 and TP2 by Gender",
       x = "Time Points",
       y = "Waist-Hip Ratio (WHR)") +
  theme_minimal()

whr_change_gender
```

### Querying self-rated health
```{r}
# Querying self-rated health scores in tp1
query_SRHealth_tp1 <- "
  SELECT shapesg_id, 
         sr_health1 AS sr_health_tp1
  FROM tp1
"
SRHealth_scores_tp1 <- dbGetQuery(conn, query_SRHealth_tp1)

# Querying self-rated health scores in tp2
query_SRHealth_tp2 <- "
  SELECT shapesg_id, 
         sr_health2 AS sr_health_tp2
  FROM tp2
"
SRHealth_scores_tp2 <- dbGetQuery(conn, query_SRHealth_tp2)

# Change in self-rated health scores between tp1 and tp2
query_change_SRHealth <- "
  SELECT tp1.shapesg_id, 
         tp1.sr_health_tp1,
         tp2.sr_health_tp2,
         (tp2.sr_health_tp2 - tp1.sr_health_tp1) AS change_in_sr_health
  FROM (
      SELECT shapesg_id, 
             sr_health1 AS sr_health_tp1
      FROM tp1
  ) AS tp1
  JOIN (
      SELECT shapesg_id, 
             sr_health2 AS sr_health_tp2
      FROM tp2
  ) AS tp2
  ON tp1.shapesg_id = tp2.shapesg_id
"
change_SRHealth_scores <- dbGetQuery(conn, query_change_SRHealth)

print(change_SRHealth_scores)
```

```{r}
# Perform paired t-test
t_test_srh <- t.test(change_SRHealth_scores$sr_health_tp1, change_SRHealth_scores$sr_health_tp2, paired = TRUE)
t_test_srh

# Summary statistics
summary_stats_srh <- change_SRHealth_scores %>%
  summarise(
    mean_sr_health_tp1 = mean(sr_health_tp1, na.rm = TRUE),
    mean_sr_health_tp2 = mean(sr_health_tp2, na.rm = TRUE),
    mean_change_in_sr_health = mean(change_in_sr_health, na.rm = TRUE),
    sd_sr_health_tp1 = sd(sr_health_tp1, na.rm = TRUE),
    sd_sr_health_tp2 = sd(sr_health_tp2, na.rm = TRUE),
    sd_change_in_sr_health = sd(change_in_sr_health, na.rm = TRUE)
  ) |> 
  print()


```

## Further analysis and modeling
### Exploration of well-being as a multi-dimensional construct
```{r}
df_1 <- d1_clean %>%
  mutate(
    hr1_tot = rowSums(across(hr1_01:hr1_06), na.rm = TRUE),
    coping_tot = rowSums(across(cope1_01:cope1_04), na.rm = TRUE),
    cog_tot = rowSums(across(cog_iw1_01:cog_iw1_10), na.rm = TRUE),
    med_tot = rowSums(across(medical1_01:medical1_17), na.rm = TRUE)
    ) |> 
  mutate(across(-shapesg_id, as.numeric))


# Correlation matrix
correlation_matrix <- cor(df_1 |> select(mo_ca1, gds1, bmi1, whr1, cog_tot, hr1_tot, coping_tot, med_tot), use = "pairwise.complete.obs")

print(correlation_matrix)

```
### Modeling cognitive performance using linear regression 
```{r}
# Linear regression to predict MOCA scores
linear_model <- lm(mo_ca1 ~ gds1 + bmi1 + whr1 + cog_tot + hr1_tot + coping_tot, data = df_1)
summary(linear_model)
anova(linear_model)
# Check residuals
plot(linear_model)
```

```{r}
# Adding covariates
linear_model_1 <- lm(mo_ca1 ~ gds1 + bmi1 + whr1 + cog_tot + hr1_tot + coping_tot + age + gender + edu_level + med_tot, data = df_1)
summary(linear_model_1)
```

### Modeling self-rated health using linear regression 
```{r}
linear_model_srhealth <- lm(sr_health1 ~ gds1 + bmi1 + whr1 + cog_tot + hr1_tot + coping_tot 
                            + age + gender + edu_level + med_tot, data = df_1)
summary(linear_model_srhealth)
```
## Close the connection
Finally, we close the connection to the database.
```{r}
dbDisconnect(conn)
```
